# Data

## Sources

```{r}

# Import all raw data from GitHub and add column dictating the state
AK <- cbind(state = 'AK', read.csv("https://raw.githubusercontent.com/kylie-brothers/US-Water-Levels/main/Raw%20Data/AK_WL_Monthly.csv"))
CA <- cbind(state = 'CA', read.csv("https://raw.githubusercontent.com/kylie-brothers/US-Water-Levels/main/Raw%20Data/CA_WL_Monthly.csv"))
FL <- cbind(state = 'FL', read.csv("https://raw.githubusercontent.com/kylie-brothers/US-Water-Levels/main/Raw%20Data/FL_WL_Monthly.csv"))
HI <- cbind(state = 'HI', read.csv("https://raw.githubusercontent.com/kylie-brothers/US-Water-Levels/main/Raw%20Data/HI_WL_Monthly.csv"))
NY <- cbind(state = 'NY', read.csv("https://raw.githubusercontent.com/kylie-brothers/US-Water-Levels/main/Raw%20Data/NY_WL_Monthly.csv"))
TX <- cbind(state = 'TX', read.csv("https://raw.githubusercontent.com/kylie-brothers/US-Water-Levels/main/Raw%20Data/TX_WL_Monthly.csv"))

```

The National Oceanic and Atmospheric Administration (NOAA) gathers information on numerous stations across the United States to monitor tides and currents. The data is collected and recorded through sensors placed on each of the stations. We will be specifically looking at the recorded water levels for 6 different stations in different regions and states throughout the United States; the stations we will be analyzing are The Battery, NY; Virginia Key, Biscayne Bay, FL; Galveston Pier 21, TX; Monterey, CA; Prudhoe Bay, AK; and Honolulu, HI.

We collected water levels for each station from January 2020 - October 2022 on a monthly basis; it is important to note that some stations have data through part of November and December, which will be removed in the data transformation phase. These collections included all of the possible data (the base elevation that was used for collection purposes), recorded in feet, but we will specifically be looking at the Mean Tide Level (MTL). Since the data is collected monthly, the time is always 00:00 GMT. By collecting the data on a monthly basis, we will be able to get the best overall long-term view of the data, roughly 23 years. This will also ensure that the data is manageable and not overwhelming, while still being able to view long term trends and seasonality.

Currently, the data has character, number, and integer fields; the only field that will need to be transformed is the date field. Additionally, majority of the states have 274 or 275 records depending on if they have November or December included, which will be transformed to only view data from October 2022 and prior. The only states that has less than 274 records are Alaska and Texas, which will be explored further in the missing patterns sections.

## Cleaning / Transformation

Overall, the data set was fairly clean and already in a usable state. However, each station had to be downloaded from the NOAA website individually, so we had to bind all of the data sets together to create one manageable data frame. Prior to binding the data frames, we created a new column for the State. Once bound together, the only field that needed to be transformed into the correct data type was the date field; we needed to convert from character to YYYY-MM-DD data format. Lastly, we filtered the data to be only from January 2000 - October 2022.

```{r}
# create full data set with all locations
df <- rbind(AK, CA, FL, HI, NY, TX)

# determine the data types of each field
str(df)

# transform date column to date data type
df$Date <- as.Date(df$Date)
df <- df[df$Date<='2022-10-01',] # Only look at value from October and prior
```

## Missing Value Analysis

```{r}
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(forcats)
library(redav)

# determine which fields have missing values
colSums(is.na(df)) %>% sort(decreasing = TRUE)
plot_missing(df, percent = TRUE)

```

Based on above charts, we can see that there are only two variables with missing data points - Lowest Tide and Highest Tide.

Since there are many other data measurements that we can use for the analysis and to view trends, seasonality, and make comparisons. Therefore, we will choose to omit these two fields from our analysis and use the other data measurements.

### Data with Missing Months

```{r}
library(lubridate)
ggplot(df, aes(x = year(ymd(Date)), fill = state)) +
  geom_bar()+
  scale_x_continuous(breaks=seq(2000,2022,by=2)) +
  scale_y_continuous(breaks=seq(0,80,by=6)) +
  labs(x = "Year", y = "Count", title = "Missing Dates by Year") +
  theme_minimal()
```

Since we have 6 stations and should only have 1 observation per month per station, we should expect to see 72 observations per year. By looking at the above chart of a count of all data by year, we can see that there are a few months missing. By data counts, mentioned earlier, we know that Alaska and Texas are the stations causing the missing values. Those are explored further and broken down by month.

#### Alaska

```{r}
# explore missing dates in Alaska
AK_missing <- df %>% filter(state=="AK")
#All years
AK_all <- ggplot(AK_missing, aes(x = year(ymd(Date)))) +
  geom_bar()+
  scale_x_continuous(breaks=seq(2000,2022,by=4)) +
  scale_y_continuous(breaks=seq(1,13,by=1)) +
  labs(x = "Year", y = "Count", title = "AK Missing Dates by Year") +
  theme_minimal()

# 2013
AK_missing_2013 <- AK_missing %>% filter(year(ymd(Date))==2013)
AK_13 <- ggplot(AK_missing_2013, aes(x = month(ymd(Date)))) +
  geom_bar()+
  scale_x_continuous(breaks=seq(1,12,by=1)) +
  scale_y_continuous(breaks=seq(0,1,by=1)) +
  labs(x = "Month", y = "Count", title = "AK Missing Dates in 2013") +
  theme_minimal()

# 2016
AK_missing_2016 <- AK_missing %>% filter(year(ymd(Date))==2016)
AK_16 <- ggplot(AK_missing_2016, aes(x = month(ymd(Date)))) +
  geom_bar()+
  scale_x_continuous(breaks=seq(1,12,by=1)) +
  scale_y_continuous(breaks=seq(0,1,by=1)) +
  labs(x = "Month", y = "Count", title = "AK Missing Dates in 2016") +
  theme_minimal()

# 2022
AK_missing_2022 <- AK_missing %>% filter(year(ymd(Date))==2022)
AK_22 <- ggplot(AK_missing_2022, aes(x = month(ymd(Date)))) +
  geom_bar()+
  scale_x_continuous(breaks=seq(1,12,by=1)) +
  scale_y_continuous(breaks=seq(0,1,by=1)) +
  labs(x = "Month", y = "Count", title = "AK Missing Dates in 2022") +
  theme_minimal()

library(patchwork)
par(mfrow=c(2,2))
AK_all + AK_13 + AK_16 + AK_22
```

Since Alaska was the only state that did not have 274 observations, so we wanted to take a deeper look into what year and months did not have a recorded observations.

Based on the first chart, we can see that there are missing values in 2013 and 2016; we are not concerned with seeing only 11 values in 2022 because we do not have that record yet.

To dig into which exact months are missing, we broke the data down further into just 2013 and just 2016. When we did this, we could see that we are missing August 2013, September 2013, and June 2016.


#### Texas

```{r}
TX_missing <- df %>% filter(state=="TX")
#All years
TX_all <- ggplot(TX_missing, aes(x = year(ymd(Date)))) +
  geom_bar()+
  scale_x_continuous(breaks=seq(2000,2022,by=4)) +
  scale_y_continuous(breaks=seq(1,13,by=1)) +
  labs(x = "Year", y = "Count", title = "TX Missing Dates by Year") +
  theme_minimal()

# 2008
TX_missing_2008 <- TX_missing %>% filter(year(ymd(Date))==2008)
TX_08 <- ggplot(TX_missing_2008, aes(x = month(ymd(Date)))) +
  geom_bar()+
  scale_x_continuous(breaks=seq(1,12,by=1)) +
  scale_y_continuous(breaks=seq(0,1,by=1)) +
  labs(x = "Month", y = "Count", title = "TX Missing Dates in 2008") +
  theme_minimal()

par(mfrow=c(1,2))
TX_all + TX_08
```

Doing similar analysis as with Alaska, we can see that Texas is only missing one month - August 2008.
